Reproducable Research Project#1
=================================

###Project Summary
This document is the first project in the Reproducable Research course as part of the Coursera Data Science Specialization track.

###Data
Activity monitoring data from 2012 was utilized for this project. Data was downloaded from [this source](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  
**Steps** The number of steps recorded during the five minute interval  
**Date** The date of observation  
**Interval** Numeric Identifier for 5 minute period  


### Download and Prepare Data
```{r echo = TRUE}
setwd("C:/Users/Gis/Downloads/repdata-data-activity/")
  options(scipen=999)
  library(reshape2)
  library(ggplot2)
activity <- read.csv("activity.csv")
  activity$date <- as.Date(activity$date, "%Y-%m-%d")
  activity.dc.date.sum <- dcast(melt(activity, "date", "steps"), date ~ variable, sum, na.rm = TRUE)
```

###Histogram - Number of Steps Per Day

```{r echo = TRUE}
qplot(activity.dc.date.sum$steps, bins = 30)
avg_steps <- round(mean(activity.dc.date.sum$steps))
med_steps <- round(median(activity.dc.date.sum$steps))
```

Participants performed an average of `r avg_steps` steps per day and median of `r med_steps` steps per day.

### Time Series Steps Per Interval


```{r echo = TRUE}
activity.dc.int.avg <- dcast(melt(activity, "interval", "steps"), interval ~ variable, mean, na.rm = TRUE)
ggplot(data = activity.dc.int.avg, aes( x = interval, y =  steps)) + 
  geom_line() +
  ggtitle("Average Steps Per Interval") +
  ylab("Average Steps")
```

## Max of Average Steps Per Interval

```{r echo = TRUE}
activity.dc.int.avg[which.max(activity.dc.int.avg$steps),]
```

## Filling in Missing Values
```{r echo = TRUE}
activity.na.count <- sum(is.na(activity$steps))
```

This data set contains `r activity.na.count` NAs for step count. I have chosen to fill these missing values in by taking the average step for each interval.

```{r echo = TRUE}
names(activity.dc.int.avg)[names(activity.dc.int.avg) == "steps"] <- "average_steps"
activity <- merge(activity, activity.dc.int.avg, by = "interval", all.x = TRUE)
activity[is.na(activity$steps), "steps"] <- activity[is.na(activity$steps), "average_steps"]
activity$average_steps <- NULL
```

### Histogram - Number of Steps Per Day after NA Substitution
```{r echo = TRUE}
activity.dc.date.sum <- dcast(melt(activity, "date", "steps"), date ~ variable, sum, na.rm = TRUE)
qplot(activity.dc.date.sum$steps, bins = 30)
avg_steps <- round(mean(activity.dc.date.sum$steps))
med_steps <- round(median(activity.dc.date.sum$steps))
```

After NA Substitution Participants performed an average of `r avg_steps` steps per day and median of `r med_steps` steps per day.

### Average steps per interval across weekdays and weekends.
```{r echo = TRUE}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity$weekday_type <- "Weekend"
activity[weekdays(activity$date) %in% weekdays1,"weekday_type"] <- "weekday"
activity.dc.int.wk.avg <- dcast(melt(activity, c("interval", "weekday_type"), "steps"), interval + weekday_type ~ variable, mean, na.rm = TRUE)
ggplot(data = activity.dc.int.wk.avg, aes( x = interval, y =  steps)) + 
  geom_line() +
  ggtitle("Average Steps Per Interval") +
  ylab("Average Steps") +
  facet_wrap(~weekday_type, nrow = 2)```
